---
title: "Single-Cell RNA-seq Analysis: From Raw Counts to Cell Communication"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi = 150
)
```

---

# Before You Start

## What this notebook teaches

This notebook walks through a complete single-cell RNA-seq (scRNA-seq) analysis of mouse
bone marrow immune cells at two time points: **Day 0** and **Day 14** (GEO: GSE142564).

Every code chunk is paired with an explanation of **why** that step exists — not just how
to run it. By the end, you will have performed:

- Multi-metric quality control and cell filtering
- SCTransform normalisation with cell-cycle regression
- Anchor-based data integration to remove batch effects
- UMAP dimensionality reduction and resolution-stable clustering
- Cluster annotation using canonical immune lineage markers
- T-cell subclustering and subtype identification
- Pseudotime trajectory inference with Monocle3
- Cell–cell communication network comparison with CellChat

## Data requirement

Download GSE142564 from NCBI GEO:
<https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE142564>

After extracting, update the paths in Section 2 to point to your local copy.

## Conceptual foundation

> **Why not just use bulk RNA-seq?**
>
> Bulk RNA-seq measures the *average* expression across all cells in a sample. In a bone
> marrow sample containing T cells, B cells, macrophages, and progenitors, the average
> signal is biologically meaningless — it represents no real cell. scRNA-seq resolves each
> cell individually, revealing the heterogeneity that bulk methods collapse away. Rare
> populations (circulating tumour cells, stem cells, regulatory T cells) that are <1% of
> the sample are completely invisible to bulk RNA-seq but detectable with scRNA-seq.

---

# Environment Setup

## Why these packages?

Each package is selected for a specific role in the pipeline. The most important ones:

- **Seurat** — the central analysis framework. Handles data loading, QC, normalisation,
  integration, clustering, and visualisation. Think of the Seurat object as a filing cabinet
  that stores every version of your data and every result in one place.
- **Matrix** — sparse matrix support. A 20,000-gene × 10,000-cell count matrix has >90%
  zeros. Storing it densely would require ~1.6 GB; sparse format needs ~50 MB. Without this,
  most laptops cannot load the data at all.
- **tidyverse** — `dplyr` for metadata manipulation, `ggplot2` for QC visualisation,
  `stringr` for the critical mouse gene name conversion in cell cycle scoring.
- **monocle3** — trajectory inference. Orders cells along a developmental path using a
  principal graph fit through the UMAP embedding.
- **CellChat** — cell–cell communication. Infers ligand–receptor interaction probabilities
  from expression data and the CellChatDB.mouse reference database.

```{r load-libraries}
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
library(clustree)   # For clustering resolution stability visualisation
```

```{r load-trajectory-libs, eval=FALSE}
# Load trajectory and communication packages when needed (Sections 9 and 10)
# Loaded here with eval=FALSE to avoid startup conflicts with Seurat
library(monocle3)
library(SeuratWrappers)
library(CellChat)
library(patchwork)
library(ComplexHeatmap)
```

---

# Data Loading and Initial Object Creation — Day 0

## The 10X Genomics count matrix

The 10X Genomics Chromium platform encapsulates single cells with barcoded gel beads in
oil droplets (GEMs). Inside each GEM, the cell's mRNA is captured and tagged with:

- A **16-nucleotide cell barcode** — unique to that gel bead, identifies the cell
- A **12-nucleotide UMI** (Unique Molecular Identifier) — unique to each captured transcript,
  used to collapse PCR duplicates

CellRanger processes raw FASTQ reads and outputs three files per sample:

| File | Contents | Represents |
|---|---|---|
| `barcodes.tsv.gz` | Cell barcode list | Column labels (cells) |
| `features.tsv.gz` | Gene ID + name list | Row labels (genes) |
| `matrix.mtx.gz` | Sparse UMI counts | Count matrix values |

`Read10X()` reads all three simultaneously and returns a sparse matrix.

> **Key insight:** The count matrix contains *UMIs*, not reads. Each UMI represents one
> captured mRNA molecule. PCR duplicates (multiple reads from the same molecule) have already
> been collapsed by CellRanger using the UMI sequence. This is why `nCount_RNA` in Seurat
> is called UMI counts, not read counts.

```{r load-day0}
# ── UPDATE THIS PATH TO YOUR LOCAL DATA LOCATION ────────────────────────────────
day0_path <- "/path/to/GSE142564_RAW/Day0/"
# ────────────────────────────────────────────────────────────────────────────────

# Read10X() reads barcodes.tsv.gz + features.tsv.gz + matrix.mtx.gz simultaneously
# Returns a dgCMatrix (sparse matrix): genes × cells
data <- Read10X(day0_path)

# Inspect dimensions before creating Seurat object
cat("Matrix dimensions (genes × cells):", dim(data), "\n")
cat("Total non-zero entries:", nnzero(data), "\n")
cat("Sparsity:", round(100 * (1 - nnzero(data) / prod(dim(data))), 1), "%\n")
```

```{r create-seurat-day0}
# CreateSeuratObject wraps the count matrix and automatically calculates:
#   nCount_RNA  = total UMIs per cell  (sequencing depth)
#   nFeature_RNA = number of genes detected per cell  (gene diversity)
Day0 <- CreateSeuratObject(counts = data)

# Print the Seurat object summary
Day0

# Inspect the metadata — this is the QC working table
head(Day0@meta.data)
```

```{r initial-gene-filter}
# Remove cells with fewer than 200 detected genes.
#
# WHY 200?
# An empty droplet (no cell captured) or a dead cell fragment will have very few
# detected genes — typically <100. Setting a floor of 200 removes the most obvious
# debris while being conservative enough not to accidentally remove real cells.
# This is a first pass; we will apply stricter thresholds after inspecting the data.

Day0 <- subset(Day0, subset = nFeature_RNA >= 200)
cat("Cells remaining after initial filter:", ncol(Day0), "\n")
```

---

# Quality Control — Day 0

## The three pillars of scRNA-seq QC

Every cell must be evaluated on three axes simultaneously:

1. **nCount_RNA** (UMI count) — sequencing depth. Too low = empty droplet or debris.
   Too high = likely doublet (two cells in one droplet).
2. **nFeature_RNA** (gene count) — transcriptional diversity. Too low = empty droplet,
   red blood cell, or debris. Too high = likely doublet.
3. **mitoRatio** (mitochondrial fraction) — cell health. High mitochondrial RNA indicates
   the cell membrane has ruptured, allowing cytoplasmic RNA to leak out while mitochondria-
   enclosed RNA is retained. These are dying cells.

A fourth metric — **log10GenesPerUMI** — captures transcriptional complexity: how many
unique genes are detected per total UMI count. Healthy, complex cells have high diversity.
Empty droplets and red blood cells have low diversity (very few genes, many copies).

## Mitochondrial fraction

```{r calc-mito}
# PercentageFeatureSet() sums UMIs from genes matching the pattern
# and divides by total UMIs per cell, returning a percentage.
#
# Mouse mitochondrial genes are named with lowercase prefix: mt-Nd1, mt-Co1, etc.
# Human mitochondrial genes use uppercase: MT-ND1, MT-CO1, etc.
# Using the wrong prefix will silently return 0% for all cells — always verify!

Day0$mitoRatio <- PercentageFeatureSet(object = Day0, pattern = "^mt-") / 100

# Sanity check: verify mitochondrial genes are present in the dataset
mito_genes_found <- grep("^mt-", rownames(Day0), value = TRUE)
cat("Mitochondrial genes found:", length(mito_genes_found), "\n")
cat("Examples:", head(mito_genes_found, 5), "\n")

# If this prints 0 genes found, check whether your data uses ^MT- instead
```

## Complexity score

```{r calc-complexity}
# log10GenesPerUMI = log10(nFeature_RNA) / log10(nCount_RNA)
#
# This is a measure of transcriptional complexity (how many unique genes
# a cell expresses relative to its total transcript count).
#
# INTUITION:
# A cell expressing 1,000 unique genes across 5,000 total UMIs has high diversity.
# A cell expressing 50 unique genes across 5,000 UMIs (e.g., a red blood cell
# almost entirely expressing haemoglobin genes) has very low diversity.
#
# The log10/log10 form compresses the scale — values typically range from 0.7–1.0
# for healthy cells. Empty droplets and RBCs cluster well below 0.82.

Day0$log10GenesPerUMI <- log10(Day0$nFeature_RNA) / log10(Day0$nCount_RNA)

# Quick summary of all QC metrics
summary(Day0@meta.data[, c("nCount_RNA", "nFeature_RNA", "mitoRatio", "log10GenesPerUMI")])
```

## Visualise QC metrics BEFORE filtering

> **This step is non-negotiable.**
> You cannot choose appropriate thresholds without first seeing the distribution.
> Thresholds from a paper or tutorial are for *their* data — your dataset may have
> a completely different sequencing depth or mitochondrial profile. Always look first.

```{r qc-violin-prefilter, fig.height=5}
VlnPlot(
  Day0,
  features = c("nFeature_RNA", "nCount_RNA", "mitoRatio"),
  ncol  = 3,
  layer = "counts",
  pt.size = 0
) +
  plot_annotation(title = "QC Metrics — Day 0 (before filtering)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

## Metadata organisation

```{r organise-metadata}
# Extract the metadata dataframe, clean column names, and store a record
# of the original counts before any filtering modifies them.

metadata <- Day0@meta.data

# Store cell barcodes as a column (they are row names by default, which is
# not accessible after dplyr operations)
metadata$cells <- rownames(metadata)

# Label this sample
metadata$sample <- "Day0"

# Preserve original values before renaming (useful for auditing)
metadata$orig_counts   <- metadata$nCount_RNA
metadata$orig_features <- metadata$nFeature_RNA

# Rename to more readable column names
metadata <- metadata %>%
  dplyr::rename(nUMI  = nCount_RNA,
                nGene = nFeature_RNA)

# Write back into the Seurat object
Day0@meta.data <- metadata
```

## Round 1 filtering — remove doublets and dead cells

```{r filter-round1}
# Apply upper-bound filters to remove:
#   - nFeature_RNA < 4000 : removes likely doublets (two cells = ~2× genes)
#   - nCount_RNA < 20000  : removes cells with anomalously high sequencing depth
#   - mitoRatio < 0.05    : removes cells with >5% mitochondrial reads (dying cells)
#
# NOTE ON THRESHOLDS:
# These values were determined by inspecting the distributions above.
# For your own dataset, adjust these based on what you observe.
# A common mistake is using someone else's thresholds without looking at your data.

before <- ncol(Day0)

Day0 <- subset(
  Day0,
  nFeature_RNA < 4000 &
    nCount_RNA  < 20000 &
    mitoRatio   < 0.05
)

cat("Cells before Round 1 filter:", before, "\n")
cat("Cells after  Round 1 filter:", ncol(Day0), "\n")
cat("Removed:", before - ncol(Day0), "cells\n")
```

## Detailed QC distribution plots

```{r qc-umi-density, fig.height=4}
# UMI (sequencing depth) distribution
# Plotted on log10 scale because the distribution spans several orders of magnitude.
# The dashed vertical line at 2000 is a suggested lower threshold reference.
# The final lower threshold used is nUMI > 3200 (applied in Round 2).

metadata <- Day0@meta.data  # refresh after filtering

metadata %>%
  ggplot(aes(x = nUMI, fill = sample, color = sample)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_x_log10() +
  geom_vline(xintercept = 2000, linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "Transcript Counts per Cell (Day 0)",
       x = "nUMI (log10 scale)", y = "Cell Density")
```

```{r qc-gene-density, fig.height=4}
# Gene count distribution
# The dashed line at 1000 is a reference; final threshold is nGene > 900.
# Cells far left of the main peak are likely empty droplets or debris.

metadata %>%
  ggplot(aes(x = nGene, fill = sample, color = sample)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "Gene Counts per Cell (Day 0)",
       x = "nGene (log10 scale)", y = "Cell Density")
```

```{r qc-mito-density, fig.height=4}
# Mitochondrial ratio distribution
# A healthy immune cell population shows >95% of cells below 0.05 (5%).
# Any cells far to the right of the main peak are dying or damaged.

metadata %>%
  ggplot(aes(x = mitoRatio, fill = sample, color = sample)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "Mitochondrial Ratio Distribution (Day 0)",
       x = "Mitochondrial Ratio", y = "Cell Density")
```

```{r qc-scatter, fig.height=5}
# The most powerful single QC figure: UMI vs. Gene coloured by mitochondrial ratio.
#
# WHAT TO LOOK FOR:
# - Healthy cells: tight linear cloud in the upper-right, coloured light grey (low mito)
# - Dead cells: lower-left region, coloured dark (high mito) — low UMI, low genes, high mito
# - Doublets: upper-right outliers beyond the main cloud — unusually high UMI AND gene counts
# - Low-complexity cells (RBCs, empty drops): points BELOW the trend line — high UMI, low genes
#
# The red lines mark the Round 2 thresholds: nUMI > 3200 (vertical), nGene > 1000 (horizontal)

metadata %>%
  ggplot(aes(x = nUMI, y = nGene, color = mitoRatio)) +
  geom_point(alpha = 0.6, size = 0.8) +
  stat_smooth(method = "lm", color = "blue", se = TRUE, linetype = "dashed") +
  scale_color_gradient(low = "gray90", high = "black") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = 3200, linetype = "dotted", color = "red", linewidth = 0.9) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "red", linewidth = 0.9) +
  facet_wrap(~sample) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "UMI vs. Gene Counts Coloured by Mitochondrial Ratio",
       x = "nUMI", y = "nGene", color = "Mito Ratio")
```

```{r qc-complexity-density, fig.height=4}
# Complexity score (log10GenesPerUMI) distribution
# The threshold 0.82 removes cells below this complexity floor.
# Red blood cells typically score ~0.6–0.75; healthy immune cells score >0.82.

metadata %>%
  ggplot(aes(x = log10GenesPerUMI, fill = sample, color = sample)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  geom_vline(xintercept = 0.82, linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Set1") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 14) +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "Transcriptional Complexity Distribution (Day 0)",
       x = "log10(Genes per UMI)", y = "Cell Density")
```

## Round 2 filtering — remove low-complexity and low-depth cells

```{r filter-round2}
# Apply lower-bound filters:
#   log10GenesPerUMI > 0.82 : complexity floor (removes RBCs, empty droplets)
#   nGene > 900             : minimum gene diversity
#   nUMI > 3200             : minimum sequencing depth
#
# WHY TWO ROUNDS?
# Round 1 removes obvious outliers at the top (doublets, dead cells).
# Round 2, enabled by the diagnostic plots above, removes the subtler
# low-quality cells at the bottom that passed the initial 200-gene floor.

before <- ncol(Day0)

Day0 <- subset(
  x = Day0,
  subset = (log10GenesPerUMI > 0.82) & (nGene > 900) & (nUMI > 3200)
)

cat("Cells before Round 2 filter:", before, "\n")
cat("Cells after  Round 2 filter:", ncol(Day0), "\n")
cat("Removed:", before - ncol(Day0), "cells\n")

# Final QC check — distributions should now be clean and unimodal
VlnPlot(Day0, features = c("nGene", "nUMI", "mitoRatio"), pt.size = 0) +
  plot_annotation(title = "QC Metrics — Day 0 (AFTER filtering)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

# Quality Control — Day 14

> **Why repeat identical QC for every sample?**
>
> Batch effects, reagent variability, and sample handling differences mean that
> the QC distributions are rarely identical across samples. Applying thresholds
> globally (e.g., from one sample's distribution) can under-filter one sample
> and over-filter another, introducing artificial transcriptional differences
> between conditions *before* any biology has been measured.
>
> Always QC each sample independently and compare the cell counts and metric
> distributions across samples before merging.

```{r load-day14}
# ── UPDATE THIS PATH TO YOUR LOCAL DATA LOCATION ────────────────────────────────
day14_path <- "/path/to/GSE142564_RAW/Day14/"
# ────────────────────────────────────────────────────────────────────────────────

data_14 <- Read10X(day14_path)
cat("Day14 matrix dimensions:", dim(data_14), "\n")

Day14 <- CreateSeuratObject(counts = data_14)
Day14 <- subset(Day14, subset = nFeature_RNA >= 200)
cat("Cells after initial filter:", ncol(Day14), "\n")
```

```{r qc-metrics-day14}
# Calculate QC metrics — identical to Day0
Day14$mitoRatio <- PercentageFeatureSet(object = Day14, pattern = "^mt-") / 100

# Verify mitochondrial genes are present (same check as Day0)
mito_day14 <- grep("^mt-", rownames(Day14), value = TRUE)
cat("Mitochondrial genes in Day14:", length(mito_day14), "\n")

Day14$log10GenesPerUMI <- log10(Day14$nFeature_RNA) / log10(Day14$nCount_RNA)

summary(Day14@meta.data[, c("nCount_RNA", "nFeature_RNA", "mitoRatio", "log10GenesPerUMI")])
```

```{r qc-violin-day14-pre, fig.height=5}
# ALWAYS inspect the Day14 distribution before applying any thresholds.
# Compare to the Day0 violin plots above — do they look similar?
# Large differences in nUMI or nGene distributions between samples
# may indicate a batch effect or sample quality difference.

VlnPlot(
  Day14,
  features = c("nFeature_RNA", "nCount_RNA", "mitoRatio"),
  ncol = 3,
  layer = "counts",
  pt.size = 0
) +
  plot_annotation(title = "QC Metrics — Day 14 (before filtering)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

```{r organise-metadata-day14}
# Identical metadata organisation as Day0
metadata_14 <- Day14@meta.data
metadata_14$cells          <- rownames(metadata_14)
metadata_14$sample         <- "Day14"
metadata_14$orig_counts    <- metadata_14$nCount_RNA
metadata_14$orig_features  <- metadata_14$nFeature_RNA

metadata_14 <- metadata_14 %>%
  dplyr::rename(nUMI  = nCount_RNA,
                nGene = nFeature_RNA)

Day14@meta.data <- metadata_14
```

```{r qc-scatter-day14, fig.height=5}
# The UMI vs. Gene scatter for Day14 — compare the cloud shape to Day0.
# Both samples should show similar linear relationships if sample quality is consistent.
# Red lines are the same thresholds as Day0 (adjust if Day14's distribution differs).

metadata_14 %>%
  ggplot(aes(x = nUMI, y = nGene, color = mitoRatio)) +
  geom_point(alpha = 0.6, size = 0.8) +
  stat_smooth(method = "lm", color = "blue", se = TRUE, linetype = "dashed") +
  scale_color_gradient(low = "gray90", high = "black") +
  scale_x_log10() + scale_y_log10() +
  geom_vline(xintercept = 3200, linetype = "dotted", color = "red", linewidth = 0.9) +
  geom_hline(yintercept = 1000, linetype = "dotted", color = "red", linewidth = 0.9) +
  facet_wrap(~sample) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "right", plot.title = element_text(hjust = 0.5, face = "bold")) +
  labs(title = "UMI vs. Gene Counts Coloured by Mitochondrial Ratio — Day 14",
       x = "nUMI", y = "nGene", color = "Mito Ratio")
```

```{r filter-day14}
# Round 1: Remove doublets and dead cells
before_14 <- ncol(Day14)
Day14 <- subset(Day14, nFeature_RNA < 4000 & nCount_RNA < 20000 & mitoRatio < 0.05)
cat("After Round 1 — removed:", before_14 - ncol(Day14), "cells, remaining:", ncol(Day14), "\n")

# Round 2: Remove low-complexity and low-depth cells
before_14b <- ncol(Day14)
Day14 <- subset(x = Day14, subset = (log10GenesPerUMI > 0.82) & (nGene > 900) & (nUMI > 3200))
cat("After Round 2 — removed:", before_14b - ncol(Day14), "cells, remaining:", ncol(Day14), "\n")
```

```{r qc-violin-day14-post, fig.height=5}
VlnPlot(Day14, features = c("nGene", "nUMI", "mitoRatio"), pt.size = 0) +
  plot_annotation(title = "QC Metrics — Day 14 (AFTER filtering)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

```{r qc-comparison}
# Compare final cell counts and QC distributions between samples
# This is the first inter-sample quality check

cat("\n=== QC Summary ===\n")
cat("Day0  — cells:", ncol(Day0),  " | median nUMI:", median(Day0$nUMI),
    " | median nGene:", median(Day0$nGene), "\n")
cat("Day14 — cells:", ncol(Day14), " | median nUMI:", median(Day14$nUMI),
    " | median nGene:", median(Day14$nGene), "\n")

# Large differences in cell count or depth between samples are a warning sign
# of a potential batch effect that will require correction in the integration step.
```

---

# Merging, Gene Filtering, and BCR/TCR Removal

## Merging both samples

```{r rename-and-merge}
# RenameCells adds a prefix to every cell barcode.
# WHY IS THIS NECESSARY?
# Cell barcodes from 10X are not globally unique — the same 16-nt sequence
# can appear in both Day0 and Day14. Without prefixing, merge() would either
# deduplicate the barcodes (losing cells) or throw an error.
# After renaming: "ACGTACGT-1" becomes "Day0_ACGTACGT-1" or "Day14_ACGTACGT-1"

Day0  <- RenameCells(Day0,  add.cell.id = "Day0")
Day14 <- RenameCells(Day14, add.cell.id = "Day14")

# merge() concatenates two Seurat objects.
# IMPORTANT: merge() is NOT integration.
# After merging, if you run PCA, cells will still cluster by sample (batch) not biology.
# Integration (Section 7) is required to correct for this.

merged <- merge(x = Day0, y = c(Day14))

cat("Total cells after merging:", ncol(merged), "\n")
cat("Cells per sample:\n")
print(table(merged$sample))
```

## Remove BCR and TCR variable genes

```{r remove-bcr-tcr}
# In a bone marrow immune dataset, immunoglobulin (Ig) variable genes and
# T-cell receptor (TCR) genes will be among the most variable genes.
#
# WHY IS THIS A PROBLEM?
# Each B cell expresses a unique V(D)J recombination product (e.g., Ighv1-26).
# Each T cell expresses unique TCR chains. These are clonotype-specific, not
# transcriptional-state-specific. If left in, PCA and UMAP will group cells
# by their clonotype (receptor sequence) rather than their functional state.
# This makes it appear that you have many more biologically distinct populations
# than actually exist, and hides real transcriptional differences.
#
# REGEX PATTERNS:
#   ^Ig[hkl]v  — Immunoglobulin heavy/kappa/lambda variable genes
#   ^Ig[hkl]j  — Immunoglobulin joining genes
#   ^Ig[kl]c   — Immunoglobulin constant genes (kappa and lambda)
#   ^Igh[adegm]— Heavy chain isotype constants (IgA, IgD, IgE, IgG, IgM)
#   ^Tr[abdg][vjc] — TCR alpha/beta/delta/gamma variable/joining/constant

before_genes <- nrow(merged)

merged <- merged[!grepl("^Ig[hkl]v",   rownames(merged), ignore.case = FALSE), ]
merged <- merged[!grepl("^Ig[hkl]j",   rownames(merged), ignore.case = FALSE), ]
merged <- merged[!grepl("^Ig[kl]c",    rownames(merged), ignore.case = FALSE), ]
merged <- merged[!grepl("^Igh[adegm]", rownames(merged), ignore.case = FALSE), ]
merged <- merged[!grepl("^Tr[abdg][vjc]", rownames(merged), ignore.case = FALSE), ]

cat("Genes before BCR/TCR removal:", before_genes, "\n")
cat("Genes after  BCR/TCR removal:", nrow(merged), "\n")
cat("Removed:", before_genes - nrow(merged), "BCR/TCR genes\n")
```

## Remove sparsely expressed genes

```{r filter-sparse-genes}
# Genes expressed in <10 cells contribute primarily noise to variable gene
# selection and PCA. Removing them:
#   1. Reduces dimensionality (faster computation)
#   2. Improves signal-to-noise ratio
#   3. Removes features that are too sparse for meaningful statistical testing

# Seurat v5 stores counts in layers; JoinLayers collapses them into one matrix
merged[["RNA"]] <- JoinLayers(merged[["RNA"]])

# Extract the count matrix
counts <- LayerData(merged, assay = "RNA", layer = "counts")

# Create a logical (TRUE/FALSE) matrix: is this gene expressed in this cell?
nonzero <- counts > 0

# Count how many cells express each gene
cells_per_gene <- Matrix::rowSums(nonzero)

# Keep only genes expressed in at least 10 cells
keep_genes <- cells_per_gene >= 10

cat("Genes before sparse filter:", nrow(merged), "\n")
cat("Genes expressed in ≥10 cells:", sum(keep_genes), "\n")
cat("Genes removed:", sum(!keep_genes), "\n")

# Subset the count matrix and rebuild the Seurat object
# (CreateSeuratObject preserves metadata via the meta.data argument)
dual_counts <- counts[keep_genes, ]
merged <- CreateSeuratObject(dual_counts, meta.data = merged@meta.data)

cat("Final merged object:", nrow(merged), "genes ×", ncol(merged), "cells\n")
```

---

# Cell Cycle Scoring

## Why cell cycle matters

Proliferating cells (in S phase or G2/M phase) have a distinct transcriptional signature —
they upregulate hundreds of genes involved in DNA replication and mitosis. In a dataset
with cycling and non-cycling cells, this signature can dominate the first principal
components, causing cells to cluster by their cell cycle phase rather than their cell type.

The strategy here is to:
1. **Score** every cell for S-phase and G2M-phase gene expression
2. **Regress** these scores out as covariates in SCTransform (next section)

This removes cell cycle as a source of variation while preserving the genuine
transcriptional differences between cell types.

```{r cell-cycle-scoring}
# A preliminary log-normalisation is required before CellCycleScoring()
# because the scoring uses gene expression levels, not raw counts.
# This normalisation is temporary — SCTransform replaces it in Section 7.

merged <- NormalizeData(merged)

# CRITICAL MOUSE/HUMAN CONVERSION:
# Seurat's built-in cc.genes list uses HUMAN gene symbols in ALL CAPS (e.g., PCNA, TOP2A).
# Mouse gene symbols use Title Case (e.g., Pcna, Top2a).
# str_to_title() converts: "PCNA" → "Pcna", "TOP2A" → "Top2a"
# Without this conversion, scoring silently returns 0 scores for ALL cells.

s.genes   <- str_to_title(cc.genes$s.genes)
g2m.genes <- str_to_title(cc.genes$g2m.genes)

cat("S-phase genes (first 5):", head(s.genes, 5), "\n")
cat("G2M-phase genes (first 5):", head(g2m.genes, 5), "\n")

merged <- CellCycleScoring(
  merged,
  g2m.features = g2m.genes,
  s.features   = s.genes,
  set.ident    = TRUE   # Set active identity to predicted phase (G1/S/G2M)
)

# Visualise canonical cell cycle marker expression across predicted phases
# These should show phase-specific enrichment if scoring worked correctly:
#   PCNA (Pcna) — DNA clamp loader, S phase
#   TOP2A (Top2a) — DNA topoisomerase, G2M phase
#   MCM6 (Mcm6) — DNA replication helicase, S phase
#   MKI67 (Mki67) — proliferation marker, G2M phase

RidgePlot(
  merged,
  features = c("Pcna", "Top2a", "Mcm6", "Mki67"),
  ncol = 2
) +
  plot_annotation(title = "Cell Cycle Marker Expression by Phase",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

```{r cell-cycle-pca, fig.height=5}
# Run a diagnostic PCA BEFORE SCTransform to see how much cell cycle drives variation.
# If cells separate by phase on PC1/PC2, regression is necessary.

merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
merged <- ScaleData(merged)
merged <- RunPCA(merged, verbose = FALSE)

DimPlot(
  merged,
  reduction = "pca",
  group.by  = "Phase",
  split.by  = "Phase"
) +
  plot_annotation(title = "PCA Coloured by Cell Cycle Phase (before regression)",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))

# If G1, S, and G2M cells separate on PC1 or PC2, the regression is working as intended.
# After SCTransform with vars.to.regress, this separation should disappear.
```

---

# SCTransform Normalisation and Data Integration

## Why SCTransform instead of log-normalisation?

Log-normalisation (dividing each cell's counts by total UMIs, then log-transforming) is
simple but imperfect. It corrects for sequencing depth but not for RNA composition effects
or the overdispersion inherent in count data.

**SCTransform** fits a regularised negative binomial regression for each gene:

```
log(E[counts]) = β₀ + β₁ × log(nUMI) + β₂ × S.Score + β₃ × G2M.Score
```

The residuals from this model are the normalised values — they represent expression
*after removing* the effect of sequencing depth and cell cycle. This is more statistically
principled because count data genuinely follows a negative binomial distribution.

```{r sctransform}
# Split the merged object back into per-sample objects before normalising.
# WHY PER SAMPLE?
# If you run SCTransform on the merged object, the model will fit a single
# depth-expression relationship across both samples simultaneously. If the
# samples have different sequencing depths (a batch effect), this conflates
# biological signal with technical variation in the normalised values.
# Running per sample first, then integrating, keeps these steps cleanly separated.

split_seurat <- SplitObject(merged, split.by = "sample")

# SCTransform is memory-intensive. This line raises the ceiling for
# objects passed to parallel worker processes. 8 GB is appropriate for
# datasets of ~5,000–20,000 cells. Scale up for larger experiments.
options(future.globals.maxSize = 8 * 1024^3)

cat("Samples to normalise:", names(split_seurat), "\n")

for (i in seq_along(split_seurat)) {
  cat("\nRunning SCTransform on sample:", names(split_seurat)[i], "\n")

  split_seurat[[i]] <- SCTransform(
    split_seurat[[i]],
    vars.to.regress = c("S.Score", "G2M.Score"),  # Regress out cell cycle
    verbose = FALSE
  )
  # Each object now has a new 'SCT' assay slot with:
  #   - scale.data : Pearson residuals (normalised, used for PCA)
  #   - data       : log1p-transformed counts (used for visualisation)
  #   - counts     : corrected counts (for downstream use)
}

cat("\nSCTransform complete.\n")
cat("Assays in object:", Assays(split_seurat[[1]]), "\n")
```

## Anchor-based integration

```{r integration}
# Seurat's integration strategy:
#
# 1. SelectIntegrationFeatures() — identifies 3000 genes that are highly variable
#    in ALL samples (not just one). These are the anchor features.
#
# 2. PrepSCTIntegration() — pre-processes the SCT residuals for anchor finding.
#
# 3. FindIntegrationAnchors() — finds "anchor" pairs: cells that are mutual
#    nearest neighbours in expression space across samples. These are cells
#    that are biologically equivalent despite coming from different samples.
#    Think of them as landmarks shared across the two datasets.
#
# 4. IntegrateData() — uses the anchor pairs to compute correction vectors and
#    project all cells into a shared embedding that removes batch-specific variation.

bm.features <- SelectIntegrationFeatures(
  object.list = split_seurat,
  nfeatures   = 3000
)

bm.list <- PrepSCTIntegration(
  object.list     = split_seurat,
  anchor.features = bm.features
)

bm.anchors <- FindIntegrationAnchors(
  object.list          = bm.list,
  normalization.method = "SCT",
  anchor.features      = bm.features
)

# MEMORY MANAGEMENT: Integration is the most RAM-intensive step.
# Remove all objects except bm.anchors before proceeding.
# On a 32 GB machine this typically requires ~20 GB of free RAM.
cat("Freeing memory before integration...\n")
remove(list = setdiff(ls(), c("bm.anchors", "split_seurat")))
gc()

bm.integrated <- IntegrateData(
  anchorset            = bm.anchors,
  normalization.method = "SCT"
)

# After integration, the 'integrated' assay contains batch-corrected values.
# Use 'integrated' for: PCA, UMAP, FindNeighbors, FindClusters
# Use 'RNA' for: gene expression quantification, FindAllMarkers, FeaturePlot values

DefaultAssay(bm.integrated) <- "integrated"

cat("Integration complete.\n")
cat("Assays available:", Assays(bm.integrated), "\n")
```

---

# Dimensionality Reduction and Clustering

## PCA and number of dimensions

```{r pca}
bm.integrated <- RunPCA(object = bm.integrated, verbose = FALSE)

# The ElbowPlot shows the standard deviation (variance explained) of each PC.
# The "elbow" — where the curve bends and flattens — indicates that additional
# PCs are capturing noise rather than signal.
# For this dataset, 20 PCs is chosen. The exact number matters less than ensuring
# you include enough PCs to capture all major cell types.

ElbowPlot(bm.integrated, ndims = 50) +
  geom_vline(xintercept = 20, linetype = "dashed", color = "red") +
  labs(title = "Elbow Plot — Variance Explained by PC",
       subtitle = "Red line = chosen dimensionality (20 PCs)")
```

## UMAP embedding

```{r umap}
# UMAP (Uniform Manifold Approximation and Projection) compresses the 20-PC
# embedding into a 2D visualisation that preserves LOCAL neighbourhood structure.
#
# KEY WARNINGS ABOUT UMAP:
#   - Distances BETWEEN clusters are NOT interpretable. Two distant clusters on the UMAP
#     may be transcriptionally similar. Two adjacent clusters may be very different.
#   - Only LOCAL structure (within-cluster and immediate neighbours) is meaningful.
#   - UMAP is for VISUALISATION only. Always use PCA space for FindNeighbors() and
#     FindClusters() — never use UMAP coordinates for these steps.

bm.integrated <- RunUMAP(
  object  = bm.integrated,
  dims    = 1:20,
  verbose = FALSE
)
```

## Graph construction and clustering

```{r clustering}
# FindNeighbors() builds a k-nearest-neighbour (KNN) graph in PCA space.
# Each cell is connected to its k most transcriptionally similar neighbours.
# This graph is the mathematical input to the community detection algorithm.

bm.integrated <- FindNeighbors(
  bm.integrated,
  reduction = "pca",
  dims      = 1:20,
  verbose   = FALSE
)

# FindClusters() applies the Leiden/Louvain community detection algorithm.
# We test 20 different resolution values (0.1 to 2.0) simultaneously.
#
# WHY TEST MULTIPLE RESOLUTIONS?
# Resolution controls the number of clusters: low resolution = few large clusters;
# high resolution = many small clusters. There is no objectively "correct" resolution —
# it depends on the biological question and the dataset.
# Testing many resolutions and visualising stability with clustree (below) allows
# an informed choice rather than an arbitrary one.

bm.integrated <- FindClusters(
  bm.integrated,
  resolution = c(seq(from = 0.1, to = 2.0, by = 0.1)),
  verbose    = FALSE
)

cat("Clustering columns added to metadata (one per resolution):\n")
print(head(grep("integrated_snn_res", colnames(bm.integrated@meta.data), value = TRUE)))
```

```{r clustree, fig.height=8}
# clustree visualises how cells move between clusters as resolution increases.
# A stable cluster at a given resolution will have a thick, consistent branch
# in the tree with no cells migrating away at small resolution changes.
#
# CHOOSING RESOLUTION:
# Select the LOWEST resolution at which biologically distinct populations emerge
# as separate clusters. Going higher gives you more clusters, but at the cost of
# splitting genuinely homogeneous populations artificially.

library(clustree)

clustree(
  x      = bm.integrated,
  prefix = "integrated_snn_res."
) +
  labs(title = "Clustering Stability Across Resolutions (20 PCs)")
```

```{r set-resolution}
# Based on the clustree output, resolution 0.1 is selected for this dataset.
# At this resolution, the main immune lineages are well separated without
# over-splitting within cell types.

Idents(bm.integrated) <- "integrated_snn_res.0.1"

cat("Number of clusters at resolution 0.1:",
    length(levels(Idents(bm.integrated))), "\n")
```

## Integration quality control

```{r umap-overall, fig.height=5}
# UMAP 1: Overall cluster assignments
DimPlot(
  bm.integrated,
  reduction = "umap",
  label     = TRUE,
  label.size = 5
) +
  NoLegend() +
  labs(title = "Integrated UMAP — Cluster Assignments (resolution 0.1)")
```

```{r umap-by-sample, fig.height=5}
# UMAP 2: Split by sample — the KEY integration QC plot.
#
# SUCCESS: Day0 and Day14 cells co-occupy the same cluster positions.
#          Each cluster contains cells from both time points.
# FAILURE: Day0 cells and Day14 cells form separate UMAP regions.
#          This means integration did not remove the batch effect.
#
# If this plot shows failure, consider: Harmony integration, increasing nfeatures,
# or checking whether the samples are biologically too different to integrate.

DimPlot(
  bm.integrated,
  reduction = "umap",
  split.by  = "sample",
  label     = TRUE
) +
  labs(title = "Integrated UMAP — Split by Sample")
```

```{r umap-by-phase, fig.height=5}
# UMAP 3: Split by cell cycle phase — confirms cell cycle regression worked.
#
# SUCCESS: G1, S, and G2M cells are distributed across all clusters (not separated).
# FAILURE: S and G2M cells form their own cluster(s), separate from G1 cells.
#          This means cell cycle regression did not fully remove the effect.

DimPlot(
  bm.integrated,
  reduction = "umap",
  split.by  = "Phase",
  label     = TRUE
) +
  labs(title = "Integrated UMAP — Split by Cell Cycle Phase")
```

---

# Cluster Annotation

## Switching to the RNA assay

```{r switch-to-rna}
# IMPORTANT ASSAY SWITCH:
# The 'integrated' assay corrects for batch effects but distorts absolute expression values.
# It is suitable for: clustering, UMAP, dimensionality reduction.
# It is NOT suitable for: differential expression, marker gene quantification, FeaturePlot.
#
# Always switch back to 'RNA' before running FindAllMarkers() or FeaturePlot().

DefaultAssay(bm.integrated) <- "RNA"

# Log-normalise the RNA assay
bm.integrated <- NormalizeData(
  object               = bm.integrated,
  scale.factor         = 10000,
  normalization.method = "LogNormalize"
)

# Scale all genes in the RNA assay
# (ScaleData is required before DoHeatmap — it is not the same as SCTransform scaling)
bm.integrated <- ScaleData(
  object   = bm.integrated,
  features = rownames(bm.integrated)
)

# Join layers (required in Seurat v5 before FindAllMarkers)
bm.integrated <- JoinLayers(bm.integrated)
```

## Canonical lineage marker expression

```{r feature-plot-markers, fig.height=7}
# These four markers cover the major immune lineages expected in bone marrow:
#   Cd3e   — Pan-T cell (part of the TCR signalling complex)
#   Cd19   — B cell (B cell receptor coreceptor)
#   Adgre1 — Macrophage (F4/80, a well-established mouse macrophage marker)
#   Cd14   — Monocyte (LPS co-receptor; lower on tissue-resident macrophages)
#
# order = TRUE ensures cells expressing the gene are plotted on top,
# making positive cells visible even in dense UMAP regions.

FeaturePlot(
  bm.integrated,
  features = c("Cd3e", "Cd19", "Adgre1", "Cd14"),
  label    = TRUE,
  ncol     = 2,
  order    = TRUE
) +
  plot_annotation(title = "Canonical Lineage Marker Expression",
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

```{r dot-plot-markers, fig.height=4}
# DotPlot encodes two dimensions of information simultaneously:
#   - Dot SIZE    = percentage of cells in that cluster expressing the gene (% expressed)
#   - Dot COLOUR  = average expression level in expressing cells (scaled expression)
#
# A cluster that is truly a B cell should show large, dark dots for Cd19
# and small, light dots for Cd3e, Adgre1, and Cd14.

DotPlot(
  bm.integrated,
  features = c("Cd3e", "Cd19", "Adgre1", "Cd14")
) +
  RotatedAxis() +
  labs(title = "Lineage Marker DotPlot — Size = % Expressing, Colour = Avg Expression")
```

## FindAllMarkers — data-driven cluster annotation

```{r find-all-markers}
# FindAllMarkers() tests each gene for differential expression in each cluster
# versus all other clusters combined (one-vs-rest).
#
# KEY PARAMETER CHOICES:
#
# logfc.threshold = 0.25
#   Pre-filters to genes with at least 0.25 log2FC difference.
#   This speeds up the test by skipping genes unlikely to be significant.
#   It does NOT mean we only care about large fold-changes — the FDR filter below
#   handles statistical significance separately from biological effect size.
#
# test.use = "negbinom"
#   Negative binomial test accounts for the overdispersion in count data.
#   scRNA-seq counts are more variable than Poisson predicts (many zeros, occasional
#   very high counts). The negative binomial model fits this better.
#   Default is Wilcoxon (faster but ignores count distribution structure).
#
# only.pos = TRUE
#   Returns only upregulated markers per cluster. Downregulated genes are less
#   useful for defining what a cluster IS (they tell you what it is NOT).
#
# return.thresh = 0.05
#   Pre-filters to FDR-adjusted p-value < 0.05 before returning results.

markers <- FindAllMarkers(
  object          = bm.integrated,
  logfc.threshold = 0.25,
  test.use        = "negbinom",
  only.pos        = TRUE,
  return.thresh   = 0.05,
  verbose         = FALSE
)

# Remove any rows where p_val_adj == 0 (a Seurat artefact in some versions
# where cells with identical expression profiles generate exact p = 0)
markers <- markers %>%
  dplyr::filter(p_val_adj > 0 & p_val_adj < 0.05)

cat("Significant markers found:", nrow(markers), "\n")
cat("Markers per cluster:\n")
print(table(markers$cluster))
```

```{r top10-heatmap, fig.height=8}
# Select top 10 markers per cluster ranked by average log2 fold-change
top10 <- markers %>%
  dplyr::group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  dplyr::slice_head(n = 10) %>%
  dplyr::ungroup()

seurat_integrated <- bm.integrated  # Alias for clarity

# DoHeatmap shows the scaled expression of top10 marker genes across all cells,
# grouped by cluster. Each column is a cell; each row is a gene.
# A well-annotated cluster should show a distinct block of high expression
# for its marker genes.

DoHeatmap(seurat_integrated, features = top10$gene) +
  NoLegend() +
  labs(title = "Top 10 Marker Genes per Cluster")
```

```{r dotplot-top-markers, fig.height=6}
DotPlot(seurat_integrated, features = unique(top10$gene)) +
  RotatedAxis() +
  labs(title = "Top Marker Gene DotPlot")
```

---

# T-cell Subclustering

## Why subcluster?

At the whole-dataset clustering resolution (0.1), the T-cell population appears as a single
cluster. However, T cells are a functionally heterogeneous group:

- **CD8+ Cytotoxic T cells** — direct cell killing
- **CD4+ Th17 cells** — inflammatory IL-17 producers
- **Regulatory T cells (Tregs)** — immune suppression via FOXP3
- **γδ T cells (IL-17a+)** — innate-like, express IL-17a and Trgv2
- **γδ T cells (IL-17a−)** — cycling/proliferating, express Top2a and Birc5

Subclustering applies the entire clustering pipeline *only to T cells*, at a finer
resolution, revealing subtypes invisible at the whole-dataset scale.

```{r t-cell-subset}
# Subset cluster 3 (identified as T cells based on Cd3e FeaturePlot above)
# IMPORTANT: Verify the T-cell cluster number in YOUR data before subsetting.
# Run: FeaturePlot(seurat_integrated, features = "Cd3e") to confirm.

T_cell <- subset(
  seurat_integrated,
  subset = integrated_snn_res.0.1 == "3"
)

cat("T cells extracted:", ncol(T_cell), "cells\n")

# Use the integrated assay for subclustering (same rationale as main clustering)
DefaultAssay(T_cell) <- "integrated"
```

```{r t-cell-clustering}
# Run the full clustering pipeline on the T-cell subset
T_cell <- RunUMAP(T_cell, dims = 1:20, verbose = FALSE)
T_cell <- FindNeighbors(T_cell, reduction = "pca", dims = 1:20, verbose = FALSE)
T_cell <- FindClusters(T_cell, resolution = c(seq(0.1, 2.0, by = 0.1)), verbose = FALSE)

# Assess resolution stability
clustree(x = T_cell, prefix = "integrated_snn_res.") +
  labs(title = "T-cell Subclustering Stability (clustree)")
```

```{r t-cell-set-resolution}
# Resolution 0.2 is selected for T-cell subclustering
Idents(T_cell) <- "integrated_snn_res.0.2"

DimPlot(T_cell, label = TRUE, label.size = 5) +
  NoLegend() +
  labs(title = "T-cell Subcluster UMAP (resolution 0.2)")
```

## T-cell subtype marker visualisation

```{r t-cell-rna-norm}
# Switch to RNA assay for marker gene quantification
DefaultAssay(T_cell) <- "RNA"
T_cell <- NormalizeData(T_cell, scale.factor = 10000, verbose = FALSE)
T_cell <- ScaleData(T_cell, features = rownames(T_cell), verbose = FALSE)
T_cell <- JoinLayers(T_cell)
```

```{r t-cell-dotplot, fig.height=6}
# Named list of marker genes per T-cell subtype
# The list structure allows DotPlot to group genes by their biological category
tcell_clusters <- list(
  Cytotoxic_T_cells        = c("Lef1", "Ms4a4b", "Cd8a"),
  Th17_cells               = c("Tnfsf8", "Cxcr6", "Cd4", "Il17a", "Rora"),
  Regulatory_T_cells       = c("Foxp3", "Ctla4"),
  Il17a_positive_gdT_cells = c("Il23r", "Trgv2"),
  Il17a_negative_gdT_cells = c("Birc5", "Top2a")
)

DotPlot(T_cell, features = tcell_clusters) +
  RotatedAxis() +
  labs(title = "T-cell Subtype Marker DotPlot",
       subtitle = "Size = % expressing | Colour = scaled expression")
```

---

# Trajectory Inference with Monocle3

## What is pseudotime?

Pseudotime is not real time. It is an abstract axis that orders cells from
transcriptionally "early" (closest to the root state) to transcriptionally "late"
(furthest from the root). The ordering is based on transcriptional similarity —
cells that are most similar to their neighbours form a continuous path.

Monocle3 learns a **principal graph** through the UMAP embedding — a mathematical
skeleton that captures the trajectory topology — and then assigns each cell a
distance along this graph from the chosen root node.

```{r trajectory-setup, eval=FALSE}
# Load trajectory packages
library(monocle3)
library(SeuratWrappers)

DefaultAssay(T_cell) <- "integrated"

# Convert Seurat → Monocle3 cell_data_set (CDS) object
# SeuratWrappers::as.cell_data_set() transfers: counts, UMAP, PCA, cluster labels
bm.integrated.cds <- as.cell_data_set(
  T_cell,
  group.by = "integrated_snn_res.0.2"
)

# Re-cluster within Monocle3 (required for learn_graph)
bm.integrated.cds <- cluster_cells(
  cds              = bm.integrated.cds,
  reduction_method = "UMAP"
)

# Learn the principal graph through the UMAP embedding
# use_partition = TRUE restricts the trajectory to each connected UMAP partition,
# preventing biologically disconnected states from being linked.
bm.integrated.cds <- learn_graph(
  bm.integrated.cds,
  use_partition = TRUE
)

# Visualise the trajectory graph overlaid on the UMAP
plot_cells(
  cds                    = bm.integrated.cds,
  show_trajectory_graph  = TRUE,
  color_cells_by         = "cluster",
  label_groups_by_cluster = TRUE
)
```

```{r pseudotime, eval=FALSE}
# order_cells() assigns pseudotime values.
# This step opens an interactive Shiny browser window — you must manually click
# to select the ROOT NODE (the biological starting state of the trajectory).
#
# HOW TO CHOOSE THE ROOT:
# 1. Look at the T-cell subtype DotPlot from Section 8.
# 2. Identify which cluster expresses the LEAST differentiated markers.
#    For T cells, naive T cells or early precursors are the logical root.
# 3. In the interactive window, click on that cluster.
#
# If running non-interactively (e.g., in a pipeline), specify the root programmatically:
#   bm.integrated.cds <- order_cells(bm.integrated.cds, root_cells = <barcode vector>)

bm.integrated.cds <- order_cells(
  bm.integrated.cds,
  reduction_method = "UMAP"
)

# Colour cells by pseudotime
# Cells close to the root (early) appear dark; cells far from root (late) appear light.
plot_cells(
  cds                   = bm.integrated.cds,
  color_cells_by        = "pseudotime",
  show_trajectory_graph = TRUE
) +
  labs(title = "T-cell Pseudotime Trajectory",
       subtitle = "Dark = early (root), Light = late (differentiated)")
```

---

# Cell–Cell Communication Analysis (CellChat)

## The biological question

Which ligand–receptor interactions occur between cell populations at Day 0 and Day 14?
Do any signalling pathways change between the two time points?

CellChat answers this by:
1. Identifying ligand–receptor (L-R) pairs where both ligand AND receptor are
   overexpressed in the relevant cell populations
2. Computing a communication probability using a mass-action model based on
   the CellChatDB.mouse database (~600 curated L-R interactions)
3. Aggregating L-R pairs to the signalling pathway level
4. Comparing the resulting networks between Day 0 and Day 14

```{r cellchat-setup, eval=FALSE}
library(CellChat)
library(patchwork)
options(stringsAsFactors = FALSE)

# Split the integrated Seurat object back into per-sample objects
split_seurat_cc <- SplitObject(bm.integrated, split.by = "sample")
split_seurat_cc <- split_seurat_cc[c("Day0", "Day14")]
```

```{r cellchat-day0, eval=FALSE}
Day0_cc <- split_seurat_cc[["Day0"]]
DefaultAssay(Day0_cc) <- "RNA"

# Input to CellChat: the log-normalised RNA layer (NOT the integrated assay)
data.input.Day0 <- LayerData(Day0_cc, assay = "RNA", layer = "data")

# Cluster labels — prefix with "group" to avoid issues with numeric cluster names
labels <- paste0("group", as.character(Idents(Day0_cc)))

# Metadata MUST have rownames matching colnames of the expression matrix
# This ordering requirement is critical — mismatches cause silent errors
metaDay0 <- data.frame(group = as.factor(labels))
rownames(metaDay0) <- colnames(data.input.Day0)

# Create CellChat object
cellchatDay0 <- createCellChat(
  object   = data.input.Day0,
  meta     = metaDay0,
  group.by = "group"
)

# Load and assign the mouse ligand-receptor interaction database
# For human data: use CellChatDB.human
CellChatDB         <- CellChatDB.mouse
cellchatDay0@DB    <- CellChatDB

# ── CORE PIPELINE ─────────────────────────────────────────────────────────────
# Step 1: Keep only signalling genes present in this dataset
cellchatDay0 <- subsetData(cellchatDay0)

# Step 2: Find overexpressed genes (above background in at least one group)
cellchatDay0 <- identifyOverExpressedGenes(cellchatDay0)

# Step 3: Find L-R pairs where BOTH ligand and receptor are overexpressed
cellchatDay0 <- identifyOverExpressedInteractions(cellchatDay0)

# Step 4: Compute communication probability (mass-action model)
cellchatDay0 <- computeCommunProb(cellchatDay0)

# Step 5: Filter out interactions from groups with < 10 cells (statistically unreliable)
cellchatDay0 <- filterCommunication(cellchatDay0, min.cells = 10)

# Step 6: Aggregate L-R pair probabilities to the signalling pathway level
cellchatDay0 <- computeCommunProbPathway(cellchatDay0)

# Step 7: Build the aggregated communication network (count matrix and weight matrix)
cellchatDay0 <- aggregateNet(cellchatDay0)

# Visualise as a circle plot: edge width ∝ interaction frequency or strength
groupSize <- as.numeric(table(cellchatDay0@idents))
par(mfrow = c(1, 2), xpd = TRUE)
netVisual_circle(cellchatDay0@net$count, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 title.name = "Number of Interactions — Day 0")
netVisual_circle(cellchatDay0@net$weight, vertex.weight = groupSize,
                 weight.scale = TRUE, label.edge = FALSE,
                 title.name = "Interaction Strength — Day 0")
```

```{r cellchat-day14, eval=FALSE}
# Identical pipeline for Day14
Day14_cc <- split_seurat_cc[["Day14"]]
DefaultAssay(Day14_cc) <- "RNA"

data.input.Day14 <- LayerData(Day14_cc, assay = "RNA", layer = "data")
labels_14 <- paste0("group", as.character(Idents(Day14_cc)))
metaDay14 <- data.frame(group = as.factor(labels_14))
rownames(metaDay14) <- colnames(data.input.Day14)
metaDay14$samples <- "Day14"  # Suppresses a CellChat warning

cellchatDay14 <- createCellChat(object = data.input.Day14, meta = metaDay14, group.by = "group")
cellchatDay14@DB <- CellChatDB.mouse

cellchatDay14 <- subsetData(cellchatDay14)
cellchatDay14 <- identifyOverExpressedGenes(cellchatDay14)
cellchatDay14 <- identifyOverExpressedInteractions(cellchatDay14)
cellchatDay14 <- computeCommunProb(cellchatDay14)
cellchatDay14 <- filterCommunication(cellchatDay14, min.cells = 10)
cellchatDay14 <- computeCommunProbPathway(cellchatDay14)
cellchatDay14 <- aggregateNet(cellchatDay14)
```

## Comparative CellChat analysis

```{r cellchat-compare, eval=FALSE}
# Compute centrality scores BEFORE merging — required for signalling role plots
cellchatDay0  <- netAnalysis_computeCentrality(cellchatDay0)
cellchatDay14 <- netAnalysis_computeCentrality(cellchatDay14)

# Merge into a single comparative object
object.list <- list(Day0 = cellchatDay0, Day14 = cellchatDay14)
cellchat    <- mergeCellChat(object.list, add.names = names(object.list))

# 1. Total interactions: has overall communication increased or decreased?
gg1 <- compareInteractions(cellchat, show.legend = FALSE, group = c(1, 2))
gg2 <- compareInteractions(cellchat, show.legend = FALSE, group = c(1, 2), measure = "weight")
gg1 + gg2 + plot_annotation(title = "Total Interactions Day0 vs Day14")

# 2. Differential interactions per cell type (which groups gained/lost communication?)
par(mfrow = c(1, 2), xpd = TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE)
netVisual_diffInteraction(cellchat, weight.scale = TRUE, measure = "weight")

# 3. Heatmap view of sender-receiver interaction matrix
gg1 <- netVisual_heatmap(cellchat)
gg2 <- netVisual_heatmap(cellchat, measure = "weight")
gg1 + gg2

# 4. Signalling role scatter: who are the dominant senders vs. receivers?
num.link <- sapply(object.list, function(x) {
  rowSums(x@net$count) + colSums(x@net$count) - diag(x@net$count)
})
weight.MinMax <- c(min(num.link), max(num.link))
gg <- lapply(seq_along(object.list), function(i) {
  netAnalysis_signalingRole_scatter(object.list[[i]],
                                    title = names(object.list)[i],
                                    weight.MinMax = weight.MinMax)
})
patchwork::wrap_plots(plots = gg)

# 5. Information flow comparison: which pathways carry the most communication weight?
# Stacked = TRUE shows absolute flow; stacked = FALSE shows change between conditions
gg1 <- rankNet(cellchat, mode = "comparison", measure = "weight",
               stacked = TRUE,  do.stat = TRUE)
gg2 <- rankNet(cellchat, mode = "comparison", measure = "weight",
               stacked = FALSE, do.stat = TRUE)
gg1 + gg2 + plot_annotation(title = "Signalling Pathway Information Flow")

# 6. Outgoing signalling patterns per cell type (who sends which pathways?)
library(ComplexHeatmap)
pathway.union <- union(object.list[[1]]@netP$pathways, object.list[[2]]@netP$pathways)
ht1 <- netAnalysis_signalingRole_heatmap(object.list[[1]], pattern = "outgoing",
                                          signaling = pathway.union,
                                          title = "Day0", width = 5, height = 6)
ht2 <- netAnalysis_signalingRole_heatmap(object.list[[2]], pattern = "outgoing",
                                          signaling = pathway.union,
                                          title = "Day14", width = 5, height = 6)
draw(ht1 + ht2, ht_gap = unit(1, "cm"))

# 7. Bubble plot: specific L-R pairs between chosen sender (4) and receivers (5:11)
netVisual_bubble(cellchat, sources.use = 4, targets.use = c(5:11),
                 comparison = c(1, 2), angle.x = 45)
```

---

# Session Information

```{r session-info}
# Always record your environment — this is essential for reproducibility.
# Anyone who wants to replicate your analysis needs to know which package
# versions were used, as Seurat and CellChat APIs change between versions.

sessionInfo()
```

---

# What to Do Next

With the results from this notebook, you have:

- An annotated UMAP of all major immune populations in bone marrow at Day 0 and Day 14
- A T-cell subtype map at single-cell resolution
- A pseudotime trajectory through T-cell developmental states
- A comparative cell–cell communication atlas between time points

**Validation steps:**
- Cross-reference top cluster markers against published databases (CellMarker 2.0, PanglaoDB)
- Use SingleR or Azimuth for automated annotation as a second opinion
- Validate one or two key L-R interactions from CellChat with co-culture experiments or spatial transcriptomics

**Extensions:**
- RNA velocity (scVelo) — adds directionality to the trajectory
- SCENIC — infers transcription factor activity per cluster
- CITE-seq — integrates surface protein data with transcriptomics
- Spatial transcriptomics — localises your clusters within tissue architecture
